---
title: "Cleaning Up a Time Series | UNHCR Refugees Package"
author: "Joe Simeone"
format: 
  html:
    embed-resources: true
    code-fold: show
    toc: true
    number-sections: true
    smooth-scroll: true 
    mainfont: inter
    theme: 
      styles.scss
editor: source

---

```{r}
#| label: "Preliminaries"
#| include: false

library(tidyverse)
library(refugees)
library(ggtext)
library(rcartocolor)
library(extrafont)


source(here::here("script", "syria_plotting_scribbles.R"))




```


# Where we're headed 

Today, I thought I'd talk a little bit about polishing graphs with `ggplot2`. Everyone here has probably made something like this before: 

```{r}
#| label: "No changes graph"
#| echo: false
#| fig-cap: "*The default themes and colors that everyone knows and loves*"

ggplot(syr_viz_dat, aes(year, totals, color = coa_tweaked)) +
  geom_line()


```


It looks, well, it probably doesn't look great. As is, it tells us... something. We've got some numbers on the y axis, 2 years on the x axis, and some countries in the legend. Otherwise, this could truly be about anything. 

And this is before we get to the aesthetics. If we were to present this output to someone, say a boss or a conference or students, they may not give our work the attention it deserves. More, they likely won't find us as credible as they otherwise could if we spent a little more time on how the graph itself *looked*.

Happily, `ggplot2` gives users a ton of flexibility and functionality for producing prettier visuals. We can break this graph (a time series) into the elements that we'll want to improve:

1. **Lines**
2. **Scales and Axes**
3. **Labels & Legends**
4. **Colors**
5. **Theming**
6. **Maybe fonts**

`Ggplot2` works by combining these elements around the `+` symbol. I'll be adding more intermediate objects usual, so you can see how these different arguments might stack together. 


::: {.callout-note}
People spend their whole careers thinking through data visualization, and I'm just a guy, so take this all in the spirit of me expressing preferences and showing some things that I find useful. 
:::

# Visuals for Exploration vs Presentation

Importantly, producing polished visuals takes **effort**, and some times a lot of effort. Because there's never enough time and we're all busy people, we want to make sure that we're spending that effort appropriately. 

We visualize data for 2 broad purposes:

1. For exploration and our own understanding
2. For presentation and others' understanding

In the exploration phase, we only care about aesthetics insofar as they facilitate our understanding of the data. This could be applying new color schemes to better see some aspect of our data, tweaking large number formatting to be more readable, or titling things to assist our recall. Here, we care much more about **speed** and **thoroughness**.

It's when we find something that we want to share that the aesthetic tools herein are most useful.

# Workflow & Getting Help

Visualization is something that requires constant trial and error. We code something, assess how our changes look, and either accept them or go back to the drawing board. The results here (hopefully) stack into a clear, linear process, but developing this required some tinkering and re-tinkering. This is mostly just to say that if you're feeling stuck or frustrated or like things are taking forever, that is often a normal part of the process. 

Moreover, because `ggplot` offers so much functionality, it is easy to forget stuff. This is true even though the syntax itself is so elegant. I cannot say enough good things about the `?` documentation. 

# The Data | United Nations High Council on Refugees (UNHCR)

UNHCR makes several of their datasets on forced migration available in the [`refugees package`](https://www.unhcr.org/refugee-statistics/insights/explainers/refugees-r-package.html). Today, we'll use the `flows` dataset to look into the aftermath of the Syrian Civil War. Specifically, we'll visualize where Syrian refugees most often sought safety from the violence. 

```{r}
#| label: "See flows data"
#| echo: true
#| collapse: true

glimpse(refugees::flows)
```

# Tidying 

Because we're interested most in visuals, I won't spend a ton of time on the data cleaning process, I'll just post what I did with a few notes. 
```{r}
#| label: "Cleaning up the data"
#| echo: true

## Narrow Data to Syria country of origin
syrian_flows <- 
  refugees::flows %>% 
  filter(coo_name == "Syrian Arab Rep." & 
           year %in% c(2011:2024)
         )

## Help re-leveling visual outcome
new_coa_levels <- 
  c("Germany", "Iraq", "Jordan", "Lebanon", "Türkiye", "Other")


## Find top 5 countries of asylum --> Pull as vector for filtering
top5_coa <- 
  syrian_flows %>% 
  summarise(sum_asy = sum(asylum_seekers),
            sum_ref = sum(refugees),
            .by = c(year, coa_name)) %>% 
  mutate(totals = sum_asy + sum_ref) %>% 
  arrange(desc(totals)) %>% 
  distinct(coa_name) %>% 
  slice_head(n = 5) %>% 
  pull()

## Filter data w/ vector, tweak a few of the variables  
syr_viz_dat <- 
  syrian_flows %>% 
  mutate(
    coa_tweaked = 
      if_else(!coa_name %in% top5_coa,
              "Other", coa_name)) %>% 
  group_by(year, coa_tweaked, .drop = FALSE) %>% 
  summarise(sum_asy = sum(asylum_seekers, na.rm = TRUE),
            sum_ref = sum(refugees, na.rm = TRUE),
            .groups = "drop"
            ) %>% 
  mutate(
    across(c(sum_asy, sum_ref), 
           ~ if_else(is.na(.x), 0, .x)
           )
  ) %>% 
  mutate(totals = 
           sum_asy + sum_ref) %>% 
  ungroup() %>% 
  mutate(coa_tweaked = as.factor(coa_tweaked),
         coa_tweaked = fct_relevel(coa_tweaked, new_coa_levels)
         )


## Code for the initial graph 
og_plot <-
  ggplot(syr_viz_dat, 
         aes(year, totals, color = coa_tweaked)) +
  geom_line()


```

Let's get after it. 

# Improving Line Asthetics 

## Fixing Turkey

So, one thing that **annoys** me about this graph is that Turkey ends prematurely. My assumption is that they stopped accepting refugees, so it's best to visualize this consistently. Here, we change our **data**, rather than our ggplot arguments. (Please excuse the janky solution). 

```{r}
#| label: "Add turkey 0s"
#| echo: true
turkey_helper <- tribble(~year, ~coa_tweaked, ~sum_asy, ~sum_ref, ~totals,
                         2019, "Türkiye", 0, 0, 0, 
                         2020, "Türkiye", 0, 0, 0, 
                         2021, "Türkiye", 0, 0, 0, 
                         2022, "Türkiye", 0, 0, 0, 
                         2023, "Türkiye", 0, 0, 0, 
                         2024, "Türkiye", 0, 0, 0)

new_coa_levels <- 
  c("Germany", "Iraq", "Jordan", "Lebanon", "Türkiye", "Other")

## Appinge data and look
syr_viz_dat <-
  syr_viz_dat %>%
  rbind(turkey_helper) %>% 
  mutate(coa_tweaked = as.factor(coa_tweaked),
         coa_tweaked = fct_relevel(coa_tweaked, new_coa_levels))

  
## Basic plot outline
plot_bones <- 
ggplot(syr_viz_dat, 
       aes(year, 
           totals,
           color = coa_tweaked, 
           fill = coa_tweaked)
       ) +
  geom_line()

plot_bones
```

## Sharpening with linewidth

Cool, now I feel like my lines could look a little sharper. This is maybe just a me thing, but this is my notebook so I make the rules. 

```{r}
plot_bones <- 
ggplot(syr_viz_dat, 
       aes(year, 
           totals,
           color = coa_tweaked, 
           fill = coa_tweaked)
       ) +
  geom_line(linewidth = .7)

plot_bones
```
We control this in the `geom_line` argument. I usually use .7 or .75, but this will depend on use case, coloring etc. You can see that it makes things a little clearer. This is also where we could control the
line style, shading etc. I'm not looking to highlight anything particular, and I don't feel like a different line style is necessary because all of our entities (countries of asylum) are the same.  

## Who doesn't love dots 
I also typically added dots to the different data points. This is especially useful when the graph is a picture. 

```{r}
#| label: "dots on dots"
#| echo: true

plot_bones <- 
ggplot(syr_viz_dat, 
       aes(year, 
           totals,
           color = coa_tweaked, 
           fill = coa_tweaked)
       ) +
  geom_line(linewidth = .7) +
  geom_point()

plot_bones
```


# Scales and Axes

Both axes could use some polish.

## X Axis | The year values are bad... 

On the X axis, we're only showing 2 years even though we have years from 2011 - 2024. This is happening because our year variable is numeric, and ggplot is making a choice based on that variable type. We can fix this by making our year variable into a date format with `lubridate::make_date`. 
```{r}
#| label: "Lubridate this baby"
#| echo: true

## Add a date variable to coerce visual - Don't care what dates are cuz we extract year 
syr_viz_dat <-
  syr_viz_dat %>%
  mutate(year_tweaked = make_date(year, 01, 01))


plot_imp_x_axis <-
ggplot(syr_viz_dat, 
       aes(year_tweaked, 
           totals,
           color = coa_tweaked, 
           fill = coa_tweaked)
       ) +
  geom_line(linewidth = .75) +
  geom_point() +
  scale_x_date(labels = scales::date_format("%Y"),
               breaks = seq(min(syr_viz_dat$year_tweaked), 
               max(syr_viz_dat$year_tweaked), 
               by = "2 year")
)

plot_imp_x_axis


```

Here, we use `scale_x_date` to control the labeling and sequencing of a variable called year_tweaked, which is variable formatted as a date value. `Labels` controls labeling. `scales::date_format` let's us pick the year value from our new date variable. `Breaks` control what values appear on the axis. We could tweak any of these to get a different result.


## Y Axis | Our numbers could like nicer

I'm actually happy with the default scale here, so we won't mess with that. On the other hand, the numbers could be a little more friendly to the eye: 

```{r}

plot_better_x_y <- 
  plot_imp_x_axis +
  scale_y_continuous(labels = scales::comma)

plot_better_x_y



```


# Coloring 

Another way to improve our plot is to move away from R's default color schemes. Because we've supplied **coa_tweaked** to the `aes()` statement, `ggplot` will map colors for both points and lines to those categories, so we just need to make a choice about the colors themselves. 

```{r}
#| label: "Some Colors"
#| echo: true

color_options <- 
rcartocolor::carto_pal(n = 12, "Bold")

# "#7F3C8D" "#11A579" "#3969AC" "#F2B701" "#E73F74" "#80BA5A" "#E68310" "#008695" "#CF1C90" "#F97B72" "#4B4B8F" "#A5AA99"
our_colors <- 
  c("#7F3C8D", "#11A579", "#3969AC", "#F2B701", "#E73F74", "#80BA5A")

plot_w_colors <- 
  plot_better_x_y + 
  scale_color_manual(values = our_colors)

plot_w_colors

```
Here, I've chosen colors from the `rcartocolor` package. I've scaled the colors manually mostly to show you this pattern for choosing colors within a pre-selected palette.


# Legends & Labels

Now, we need to annotate our graph so that people have some idea of what they're seeing. A package called `ggtext` is often handy for this purpose because it lets you use markdown syntax for your graph.

```{r}
#| label: "Make our labels"
#| echo: true
#| fig-width: 10
#| fig-height: 6

## Coloring our subtitle string
# subtitle_str <- "*Refugees Flows in the Aftermath of the Syrian Civil War | Top Receiving Countries* <br> 
#        **Türkiye**, **Jordan**, **Lebanon**, **Iraq**, **Germany**, **Everyone Else**"

countries <- c("Germany", "Iraq", "Jordan", "Lebanon", "Türkiye", "Everyone Else")

# Apply colors
colored_countries <- paste0(
  "<span style='color:", our_colors, "'><b>", countries, "</b></span>"
)

# Now reconstruct the subtitle
subtitle_str_colored <- paste0(
  "*Refugees Flows in the Aftermath of the Syrian Civil War | Top Receiving Countries* <br> ",
  paste(colored_countries, collapse = ", ")
)


syr_labs <-
  labs(title = "With the Exception of **Germany**, countries of closest proximity to Syria processed the most Syrian Refugees",
       subtitle = subtitle_str_colored,
       y = "Flows (Counts)",
       x = "",
       caption = 
       "*1. Annual counts derived from UNHCR's Refugee Package.* <br>
       *2. For techincal details, see package documentation.* <br>
       *3. Plot Made By Joe Simeone | Mistakes are his and his alone :).*")


ggtext_theme <-
  theme(plot.title = element_markdown(hjust = .5,
                                      size = 14),
        plot.subtitle = element_markdown(hjust = .5,
                                         size = 13),
        plot.caption = element_markdown(hjust= 0,
                                        size = 12),
        axis.text.y = element_markdown(),
        legend.position = "none")

plot_w_labs <- 
  plot_w_colors + 
  syr_labs + 
  ggtext_theme

plot_w_labs

```

When we render, `ggtext` converts the asterisks to bold the text, br to break lines, and additional html to add color. I've also removed the x axis label because it's redundant. Also, also, I've removed the default legend because I am the world's foremost default legend hater.

::: {.callout-note}

You'll see that I've saved the labs to their own separate object. I don't show it here, but this can be especially useful when you have to make several similar graphs depicting different groups. Specifically, it helps to facilitate functions and iteration.

:::

# Theming 

Ok, so we've dealt with all of the defaults except the graph itself. We've already gotten a taste of `theme` to add markdown elements to our labels. We can also use it to tweak additional graph aesthetics. 

```{r}
#| label: "Finished theming"
#| echo: true
#| fig-width: 10
#| fig-height: 6

## Load a font from google || Not totally prepared to talk about this one... 
sysfonts::font_add_google(name = "Source Sans Pro", family = "inter")
showtext::showtext_auto(enable = TRUE)

theme_fin <-
  ggtext_theme +
  theme(axis.line = element_line(color = "black"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "black",
                                          linetype = "dashed",
                                          linewidth = .1),
       panel.background = element_rect(fill = 'snow'),
       plot.background = element_rect(fill = "snow"),
       text = element_text(family = "inter"),
       axis.text = element_text(size = 13))


plot_w_theme <- 
  plot_w_colors + 
  theme_fin + 
  syr_labs

plot_w_theme
```
`theme` allows users to change basically any part of the graph. There are too many options to note, but the arguments typically follow the element. convention you see on the left hand side, followed by one of `element_rect()`, `element_line()`, or `element_text()`.

And that's it! Or it least that's the best I could do with the time that I had. What else would you change? 
